---
title: "Lab 4"
subtitle: "Public Health 241: Statistical Analysis of Categorical Data"
author: "YOUR NAME / YOUR STUDENT ID HERE"
date: "TODAY'S DATE"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(epiR)
library(ORCI)
library(knitr)
wcgs <- read.dta("data/wcgs.dta")
```

In this lab, we'll cover tools you'll need to complete Homework 5, including:

* Hypothesis testing
* Confidence intervals
* Obtaining critical values and p-values from Gaussian and $\chi^{2}$

# 1 Hypothesis tests, confidence intervals, and $\chi^{2}$
We’ll be using the WCGS dataset, as in lab Load it into R. In this dataset, each person in the study is represented by a row. Each row contains, among other variables, the values of two binary (0/1 indicator) variables called chd69 (Coronary Heart Disease) and arcus0 (Arcus senilis, a ring in the corneal margin or around the iris that can indicate CHD).

## 1.1 The Normal approximation for proportions, exact confidence intervals, and Fisher’s exact test
Using the Normal curve for inferences about proportions is actually an approximation. For confidence intervals for a proportion, the number of observed successes and failures should be at least 10. For hypothesis tests about a proportion, the expected numbers of successes and failures should be at least 10. For two-sample problems with binary outcomes, where the data are summarized in a 2 x 2 table, the criteria are a bit different – many authors say that the observed counts in each cell should be greater than 5 to use the Normal approximation, and for the two-sample test for proportion or the equivalent $\chi^{2}$ test, the expected counts should all be greater than 5.

What if our data don’t satisfy the assumptions for the Normal or $\chi^{2}$ distributions? For these situations, there are exact tests based on the binomial distribution and its relatives. Using `epi.2by2` from the `epiR` package, R  allows us to request exact p-values and construct "exact" confidence intervals.

### 1.1.1 Creating a 2 x 2 table
We begin by creating a 2 x 2 table, using the `table` function:

*Generic:* `table(data$dependentvariable, data$independent_variable)`


```{r}
chd.arc <- table(wcgs$chd69, wcgs$arcus0)
chd.arc
```
This creates a basic 2 x 2 table for our data. However, in order to pass this to `epi.2by2` to do the calculations, we need to have the disease/exposed count on the upper-lefthand corner and the non-disease/non-exposed count on the lower right-hand corner.  

### 1.1.2 Modifying the table to fit `epi.2by2`
Since our data is in the wrong order, we need to reorder our table before we can pass it to `epi.2by2`. To do this, we can use `relevel`. 

*Note: you only need to do this if your data is in the wrong order*

```{r}
chd.arc <- table(relevel(as.factor(wcgs$chd69), '1'), relevel(as.factor(wcgs$arcus0), '1'))
chd.arc
```

### 1.1.3 Using `epi.2by2`
We can now pass our table to `epi.2by2` to tabulate our data.

*Generic:* `epi.2by2(table)`

```{r}
epi.2by2(chd.arc)
```

### 1.1.4 Specifying the confidence interval
The default confidence level for `epi.2by2` is 95%. To override this default, reassign the `conf.level` parameter. The confidence level must be between 0 and 1.

*Generic:* `epi.2by2(table, conf.level = 0.99)`

```{r}
epi.2by2(chd.arc, conf.level = 0.99)
```

### 1.1.5 Specifying the test method
`epi.2by2` can also calculate appropriate measures of association and measures of effect in the exposed and total population based on the study design upon which the data was collected. By default, `epi.2by2` uses cohort-count. To override this default, reassign the `method` parameter. Options are cohort.count, cohort.time, case.control, or cross.sectional

*Generic:* `epi.2by2(table, method = "method")`

```{r}
epi.2by2(chd.arc, method = "case.control")
```

<!-- #### 1.1.6 Other confidence intervals -->
<!-- You can also find other confidence intervals using the `ORCI` [package](https://rdrr.io/cran/ORCI/). All functions from this package that calculate confidence intervals follow the format:  -->
<!-- > type.CI(num.events1, num.trials1, num.events2, num.trials2, conf = 0.95) -->

<!-- 1. Woolf -->
<!-- > Woolf.CI(num.events1, num.trials1, num.events2, num.trials2) -->

<!-- ```{r} -->
<!-- chd.arc -->
<!-- chd.arc.woolf.ci <- Woolf.CI(255, 3125, 941, 3152) -->
<!-- chd.arc.woolf.ci -->
<!-- ``` -->

## 1.2 Obtaining critical values and confidence levels
We learned previously how to find the p-value associated with a particular $\chi^{2}$ test statistic by using `pchisq`:  
*Remember this value is the area in the right hand tail of the $\chi^{2}$ distribution with a specified number of degrees of freedom.*
```{r}
# 1 is the number of degrees of freedom and 13.6382 is our 
# test statistic from the WCGS dataset example above 
1 - pchisq(13.6382, df = 1)    

# or, more elegantly:
pchisq(13.6382, df = 1, lower.tail = FALSE)     # recommended
```
This is because all cumulative probability functions in R compute left tail probabilities by default. To find the right tail probability, set `lower.tail = FALSE`. This is recommended over taking $1 - p$.

We can similarly find the p-value associated with a particular test statistic (z-value) based on the Standard Normal distribution. For a 2-sided test, this would look like:
```{r}
2 * (1 - pnorm(1.96))
```

### 1.2.1 Inverses
The `pchisq`, and `pnorm` functions have inverses `qchisq` and `qnorm` respectively. These functions take a probability as input and will give you its associated critical value. Be careful about making adjustments for 2-tailed tests if you want a particular confidence level. For example, for a 2-tailed test at the 95% confidence level (where $\alpha$ = 0.05), the critical z-value is found as follows:
```{r}
qnorm(.975)
```


Looking up two tails of the $\chi^2$ distribution for a hypothesis test is only appropriate when con- structing a confidence interval. Recall that the $\chi^2$ distribution is not symmetric, so it is necessary to to look up both tails separately.
```{r}
qchisq(.025, 1)           # where 1 is the degree of freedom
qchisq(.975, 1)
```

# 2 Creating and modifying variables with R

Objects are created in R using the assignment operator `<-`. For example, to create the `population.size` object that stores `1000`, type the following:
```{r}
# assigns 1000 to population.size
population.size <- 1000

# to print the value of a object, simply enter the object name:
population.size
```


To change an object value, just reassign it with the same operator. However, note that this will clear the previous value completely.
```{r}
# reassign population.size to 2000
population.size <- 2000
population.size
```


To represent a variable in R, we use a data vector, which is list of values of the same type. Think of this as a column in a dataset. To create a vector, use the command `c`.

> *Generic:* `c(value1, value2, value3 ...)


For example, let's create a vector to represent the variable age for a cohort of 5 people ages 25, 36, 35, 65, and 8:
```{r}
age <- c(25, 36, 35, 65, 8)
```

## 2.1 Conditioned variables
Perhaps the age data in your dataset is from 2014 and you now need their ages in 2019. How do we adjust the age vector? It's simple---R handles vector addition very nicely. If you add two vectors of the same length the elements will add element-wise; if you add an integer to a vector, that integer will be added to each of the elements in the vector as follows:
```{r}
age.2019 <- age + 4
age.2019
```


If you want a new variable `adult` to take the value 0 when `age` is under 21 and 1 when `age` is 21+, you can do the following using the `ifelse` function. The `ifelse()` function will evaluate the `condition` parameter and if it evaluates to `TRUE`, will return the `yes` value. Otherwise, the `condition` evaluates to `FALSE` and the function returns the `no` value. 

*Generic:* `ifelse(condition, yes, no)`


```{r}
adult <- ifelse(age >= 21, 1, 0)
adult
```


Instead of a number, say you want ages under 21 to be recoded as "child" and 21+ to be recoded as "adult". You can do so again using the `ifelse` function, but supplying strings instead of integers for the `yes` and `no` values:
```{r}
age.category <- ifelse(age >= 21, "adult", "child")
age.category
```



**For your reference: a complete list of comparison operators** *(these commands won’t work in the current environment as the variables a and b don’t exist):*
```{r eval=FALSE}
a > b     # greater than
a < b     # less than
a >= b    # greater than or equal to
a <= b    # less than or equal to
a == b    # equal to
a != b    # not equal to
```
**IMPORTANT: Note that a single equal sign `=` is similar to `<-` and will work as an object assignment, whereas the double equals `==` will compute a comparison.**



## 2.2 Boolean expressions
However, if we are just coding binary values (0/1), we don't have to use `ifelse()` at all. In fact, we can just assign the new variable based on a boolean expression, an expression that evaluates to a `TRUE` or `FALSE` value.
```{r}
# option 1:
adult1 <- age >= 21
adult1 

# option 2:
adult2 <- !(age < 21)      # ! is the negation operator
adult2
```


As you can see, there are two ways to code this, but option 1 is preferred because it allows our code to be very intuitive: the `adult` variable takes a `TRUE` (1) value when age is >= 21. 

In R, `TRUE` is represented numerically as 1 and `FALSE` as 0. Conversely, 0 evaluates to `FALSE` and all other integers evaluate to `TRUE`. Thus to truly assign binary (0/1) numeric values to the variable `adult`, we coerce the type of the condition to numeric.
```{r}
adult <- as.numeric(age >= 21)
adult
```

```{r results="hide"}
age <- c("infant", "infant", "adult", "adult", "adult", "infant")
heart.rate <- c(132, 96, 42, 64, 86, 210)
```

## 2.3 AND, OR, NOT
These boolean expressions can be very powerful with the introduction of AND, OR and NOT. For example, consider a dataset with variables `age` and `heart.rate`. To encode a new variable `healthy.hr` that is dependent on both `age` and `heart.rate`, you can type the following:
```{r}
healthy.hr <- (age == "infant" & heart.rate > 80 & heart.rate < 160) |
              (age == "adult" & heart.rate > 60 & heart.rate < 100)
healthy.hr
```





