---
title: "Lab 3"
subtitle: "Public Health 241: Statistical Analysis of Categorical Data"
author: "YOUR NAME / YOUR STUDENT ID HERE"
date: "TODAY'S DATE"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(foreign)
library(epitools)
library(epiR)
library(ORCI)
library(knitr)
library(vcd)
```

In this lab, we'll cover the following topics:

* Homework 2, in R with the `epiR` and `epitools` packages
* How to carry out a $\chi^2$ test and obtain a p-value

Many of these topics and code smippets will be useful for Homework 4.

## 1. Homework 2, Revisited

### 1.1 *Problem 1.*

You worked with this data in `lab02`, but let's briefly run through how to solve a problem like this in R. In order to recreate the table that appears on the assignment, we must learn how to create a `data.frame`, or a table.

Pay close attention to the following code:

``` {r data.frame.example}
population <- data.frame("disease"=c(140, 180), "no.disease"=c(60, 620))
population
```


Later in this lab, you will learn how to calculate all of these values (odds ratio, relative risk, etc.) through a statistical package called `epiR`, but for now, we can calculate these manually quite easily using dataframe access methods:

- The odds of disease among men:

``` {r odds.men}
prob.disease.given.man <- population$disease[1]/(population$disease[1]+population$no.disease[1])
prob.disease.given.man / (1 - prob.disease.given.man)
```

and so on. Practice doing these calculations on the rest of problem 1 (b,c,d) and problem 2.

\newpage

## 2. Cohort and Case-Control Studies using `epi.2by2`

From lecture, we have a foundational understanding of the difference between the cohort and case-control studies, which can be easily simulated in R using a function from the `epiR` package called `epi.2by2`. You should be aware of the limitations of each study design, as described in Chapter 5 of this course's textbook. Let's try this out now.

**IMPORTANT**: The `data` argument in both studies must be a `table` object in order for the function to run correctly. If you have a dataframe representation of your data, you must turn it into matrix then table. An example is shown below.

### 2.1 **Cohort Study**: 

*Generic*: `epi.2by2(data, method = "cohort.count", conf.level = 0.95, units = 100, homogeneity = "breslow.day", outcome = "as.columns")`

* Note that if you’re dealing with a cohort study’s data, the value given for Attributable Risk is bogus.

*Example from Problem 1 Assignment 2*:
``` {r cohort}
pop.matrix <- data.matrix(population)
pop.table <- as.table(pop.matrix)
pop.table
epi.2by2(pop.table, method = "cohort.count")
```
\newpage

### 2.2 **Case-Control Study**: 

*Generic*: `epi.2by2(data, method = "case.control", conf.level = 0.95, units = 100, homogeneity = "breslow.day", outcome = "as.columns")`

* Note that if you’re dealing with a case-control study’s data, the value given for Attributable Risk is an approximation that is only valid if your disease is rare (p. 50)

*Another Example from Problem 1 Assignment 2*:
``` {r case.control}
pop.matrix <- data.matrix(population)
pop.table <- as.table(pop.matrix)
pop.table
epi.2by2(pop.table, method = "case.control") # Notice the difference from above?
```

\newpage

## 3. Chi-Squared Test from Function Calls

The output of both studies shows the chi-squared value near the bottom: `165.901 p-value: < 0.001`, which is what we expect. Just to see a more reasonable example, let's pretend that we have a $\chi^2$ test statistic of 7. and we wanted to calculate its corresponding p-valie. Another option (rather than running `epi.2by2`) for finding the p-value associated with a value in a $\chi^2$ distribution is to type `pchisq()`. More specifically:

``` {r chi}
pchisq(7, df=1, lower.tail=FALSE)
```

The `pchisq` function expects the test statistic as its first entry, then a comma, then the degrees of freedom. If `lower.tail=FALSE`, then it gives us the probability to the **left** of the test statistic. If we need the **right** side, we can assign the argument to `lower.tail=TRUE`.

``` {r chi.tail}
pchisq(7, df=1, lower.tail=TRUE)
```

To lear more about the built-in statistical distributions and how to use them to get p-values, you can go to the documentation linked here: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html 


